---
title: "Problem Set: Vectors, Strings, Factors, and Dates"
format:
  html:
    toc: true
    toc-depth: 2
    code-overflow: wrap
editor: visual
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(nycflights13)
library(lubridate)
```

------------------------------------------------------------------------

## Question 1 — Exploring the Star Wars Universe

The `dplyr::starwars` dataset (available as soon as you load `tidyverse`) contains physical and biographical information on characters from the Star Wars films. Start by running `View(starwars)` and ?starwars to get oriented.

The code below stores the characters with blue eyes in the "blue_eyes" dataframe, and the characters with red eyes in the "red_eyes" dataframe. There is one character with both blue and red eyes. Use a logical vector approach again with str_detect to find out which one. Hint: one way to do it is with the logical & operator, which returns all rows for which multiple criteria are met

```{r}
# Characters with "blue" in eye_color
blue_eyes <- starwars |>
  filter(str_detect(eye_color, "blue"))
blue_eyes

# Characters with "brown" in eye_color
red_eyes <- starwars |>
  filter(str_detect(eye_color, "red"))
red_eyes

#characters with both eyes
rb_eyes <- starwars |>
  filter(str_detect(eye_color, "blue") & str_detect(eye_color, "red"))


```

## Question 2 — American Unemployment Over Time

The `ggplot2::economics` dataset contains monthly U.S. macroeconomic indicators from 1967 to 2015. The `date` column is already a proper `Date` object.

**(a)** The goal here is to see how the unemployment rate changes over time. Start by plotting the number of unemployed people over time:

```{r}
  ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line()
  
```

Now use lubridate to plot only the unemployment data from Jan 1st 2008 to Jan 1st 2009. Hint: use the lubridate function ymd to specify date windows, and use a similar logical criteria to in Q1, above

```{r}
economics |> 
  filter(date >= ymd("2008-01-01") & date <= ymd("2009-01-01"))|>
  ggplot(aes(x=date, y=unemploy))+
  geom_line()
  
```

What do you notice about the format of the date variable on the x axis of this plot vs the plot over the entire time window. What do you think is happening here?

**Your answer:** In this new graph, since the window of time is much narrower, it is possible to see details in the trend line that would otherwise be missed in a graph whose x axis covers 50 years

Now use mutate to calculate a new variable - the monthly unemployment rate as `unemploy / pop * 100`. Make a plot over the whole time period as in part 1 of this question.

```{r}
economics |>
  mutate(unemploy_rate = unemploy/pop*100) |>
  ggplot(aes(x = date, y = unemploy_rate))+
  geom_line()
```

Do you draw different conclusions now about how unemployment has changed over these 40 years? what explains the difference?

**your answer:** This graph does not illustrate the unemployment trends with time any further than part one, because the range of time (the x axis range) is too great for a simple change such as incorporating population to reveal any further information or insight. The big difference here is just that the rate reaches a maximum of 5, vs raw unemployment which reaches a population of roughly 15000

## Question 3 — Flight Status and the Value of Factors

Run the code below to create a `status` variable that categorises each flight by how early or late it arrived.

```{r}
flight_status <- flights |>
  mutate(
    status = case_when(
      is.na(arr_delay)      ~ "cancelled",
      arr_delay < -30       ~ "very early",
      arr_delay < -15       ~ "early",
      abs(arr_delay) <= 15  ~ "on time",
      arr_delay < 60        ~ "late",
      arr_delay < Inf       ~ "very late",
    ),
    .keep = "used"
  )
```

**(a)** Without converting `status` to a factor, run the followong code to create a bar chart showing the count of flights in each category. What do you notice about the ordering of the bars? Is it meaningful?

```{r}
ggplot(flight_status, aes(x = status)) +
  geom_bar()
```

**your answer:** The bars are arranged in alphabetical order, rather than an order that makes sense for what each variable actually means. This chart provides information, but it would make more sense to arrange the chart intentionally from most late to least late/early or the other way around.

**(b)** Convert `status` to a factor with levels in a natural severity order: `"very early"`, `"early"`, `"on time"`, `"late"`, `"very late"`, `"cancelled"`. Save it back into `flight_status` and recreate the bar chart. How does the chart change?

```{r}
flight_status <- flight_status |>
  mutate(status = factor(status, levels = c("very early", "early", "on time", "late", "very late", "cancelled")))

#, ordered = TRUE)

ggplot(flight_status, aes(x = status)) +
  geom_bar()
```

**your answer: This changed the order the status levels were presented so that they follow a logical progression of time. It presents what ends up appearing like a bell curve, showing most flights are on time, and fewer and fewer arrive**
